Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia

### 2024-02-08 analysis/analysis.qmd
Adjust title

---
title: Palaeodistribution of crop progenitors and other plant resources over the Pleistocene–Holocene transition in Southwest Asia
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliation: University of Bern
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: University of the Basque Country
abstract: |
  This paper presents the first comprehensive reconstructions of the prehistoric 
  ranges of [X] plant species known to have been gathered or cultivated by early
  agricultural societies in Southwest Asia, including the progenitors of the 
  first crops. We used machine learning to train an ecological niche model (ENM)
  of each species based on its present-day distribution in relation to climate 
  and environmental variables. Predictions of the potential ranges of these 
  species at key stages of the Pleistocene–Holocene transition could then be
  derived from these models using palaeoclimate 
  Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
  compared to today, indicating [...]. The modelled ranges predict the
  observed occurrence of these species on archaeological sites with [low|medium|high]
  accuracy. The regional ubiquity of species in the archaeological record is [not] 
  correlated with the predicted size of its range and the diversity of
  archaeobotanical assemblages is [not] correlated with the predicted diversity
  of its environs. This indicates that trends in taxonomic composition of
  the archaeobotanical record is [not] likely to be influenced by environmental 
  change and species turnover, not just, as is often assumed, human economic 
  choices.
format: html
bibliography: references.bib
---

```{r dependencies}
library("dplyr")
library("fs")
library("furrr")
library("ggplot2")
library("ggspatial")
library("gt")
library("here")
library("parsnip")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("rgbif")
library("rpaleoclim")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidyr")

@@

```{r tbl-occ-count}
#| tbl-cap: "Species modelled in this study."
# TODO: add to caption:
# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
# TODO: add count of species included?
flora |>
  mutate(
    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
    include = n_occ > 50
  ) |>
  select(-taxon_original) |>
  group_by(include) |>
  arrange(-n_present) |>
  gt() |>
  cols_merge(
    c(taxon, taxon_detail),
    rows = !is.na(taxon_detail),
    pattern = "{1}<br><small>(incl. {2})</small>"
  ) |>
  fmt_percent(p_present, decimals = 0) |>
  cols_merge_n_pct(n_present, p_present) |>
  tab_style(cell_text(style = "italic"), cells_body(taxon))
```

@@

```{r predict}
# TODO: tidy
rf_fit <- filter(rf_fits, taxon == "Triticum urartu") |>
  pull(fit)
rf_fit <- rf_fit[[1]]

@@

ggplot() + 
  geom_stars(data = st_crop(rf_pred, w_asia)) + 
  facet_wrap(~period) +
  scale_fill_batlow(reverse = TRUE)
```

>>>>

---
title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliation: University of Bern
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: University of the Basque Country
abstract: |
  This paper presents the first comprehensive reconstructions of the prehistoric 
  ranges of [X] plant species known to have been gathered or cultivated by early
  agricultural societies in Southwest Asia, including the progenitors of the 
  first crops. We used machine learning to train an ecological niche model (ENM)
  of each species based on its present-day distribution in relation to climate 
  and environmental variables. Predictions of the potential ranges of these 
  species at key stages of the Pleistocene–Holocene transition could then be
  derived from these models using palaeoclimate 
  Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
  compared to today, indicating [...]. The modelled ranges predict the
  observed occurrence of these species on archaeological sites with [low|medium|high]
  accuracy. The regional ubiquity of species in the archaeological record is [not] 
  correlated with the predicted size of its range and the diversity of
  archaeobotanical assemblages is [not] correlated with the predicted diversity
  of its environs. This indicates that trends in taxonomic composition of
  the archaeobotanical record is [not] likely to be influenced by environmental 
  change and species turnover, not just, as is often assumed, human economic 
  choices.
format: html
bibliography: references.bib
---

```{r dependencies}
library("dplyr")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggspatial")
library("gt")
library("here")
library("parsnip")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("rgbif")
library("rpaleoclim")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidyr")

@@

```{r tbl-occ-count}
#| tbl-cap: "Species modelled in this study."
# TODO: add to caption:
# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
# TODO: add count of species included?
flora |>
  mutate(
    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
    #include = n_occ > 50
  ) |>
  select(-taxon_original) |>
  # group_by(include) |>
  filter(n_occ > 50) |>
  arrange(-n_present) |>
  gt() |>
  cols_merge(
    c(taxon, taxon_detail),
    rows = !is.na(taxon_detail),
    pattern = "{1}<br><small>(incl. {2})</small>"
  ) |>
  fmt_percent(p_present, decimals = 0) |>
  cols_merge_n_pct(n_present, p_present) |>
  tab_style(cell_text(style = "italic"), cells_body(taxon))
```

@@

```{r predict}
# TODO: tidy
rf_fit <- filter(flora, taxon == "Triticum turgidum dicoccum") |>
  pull(fit)
rf_fit <- rf_fit[[1]]

@@

# Order and relabel periods
climate_periods <- rf_pred |>
  st_get_dimension_values("period") |>
  ordered(levels = c("LGM", "HS1", "BA", "YDS", "EH", "cur")) |>
  recode(
    "LGM" = "LGM (21 ka)",
    "HS1" = "HS1 (17.0-14.7 ka)",
    "BA"  = "BA (14.7-12.9 ka)",
    "YDS" = "YDS (12.9-11.7 ka)",
    "EH"  = "EH (11.7-8.326 ka)",
    "cur" = "current"
  )

rf_pred <- st_set_dimensions(rf_pred, "period", values = climate_periods)

ggplot() + 
  geom_stars(data = rf_pred) +
  facet_wrap(~period, ncol = 2) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = latlong) +
  labs(
    title = "Triticum turgidum dicoccum",
    fill = "Predicted occurrence",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "italic")
  )
```
